Predator clusters

### 2023-05-31 reports/04_pred_clus.qmd
Updated predator clusters from raw abundance to ind nm^-1

library(cluster)
library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_stations %>%
  as_tibble() %>% 
  group_by(amlr.station, species) %>% 
  summarize(count_norm = sum(count_norm), .groups = "drop") %>%
  group_by(species) %>%
  summarize(stations_present = n(),
            station_frac = stations_present / total_stations,
            .groups = "drop") %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

frac_fn <- scales::label_percent(accuracy = 0.1)
pred_frac %>% 
  set_names(c("Species", "Stations", "Fraction")) %>% 
  mutate(Fraction = frac_fn(Fraction)) %>% 
  knitr::kable(align = "lrr") %>% 
  kableExtra::kable_styling("striped",
                            full_width = FALSE) %>% 
  kableExtra::row_spec(cutoff_row, 
                       extra_css = "border-bottom: 2px dashed red")

@@

* Cluster 1 = marginal ice zone (SNPT)
* Cluster 2 = open ocean (ANFU, CAPT, BLPT)
* Cluster 3 = pagophilic (FUSE, ADPN, CRSE)

@@

pred_clust_sf <- tibble(amlr.station = names(clust),
                        pred_clust = factor(clust)) %>% 
  left_join(select(zoop_sf,
                   amlr.station,
                   geometry,
                   Year),
            by = "amlr.station") %>% 
  st_as_sf() %>% 
  st_transform(ant_proj())

@@

nmds_df <- sightings_nmds$points %>% 
  as_tibble(rownames = "amlr.station") %>% 
  rename_with(~ paste0("N", .x), starts_with("MDS")) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

@@

* Predator cluster 1 (marginal ice) is associated with higher phaeopigment. An ice-edge effect?
* Predator cluster 2 (open ocean) is associated with higher temperatures and chl a. Associated with macrozooplankton communities 2a and 2b (small euphausiids, amphipods, etc).
* Predator cluster 3 (pagophilic) is associated with higher salinity and deeper mixed layer. Associated with macrozooplankton communities 3a and 3b (big euphausiids, Metridia, etc).
* No clear relationship between predator community and ice categories.

@@

# Hydrology plot
ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               data = envfit_vecs,
               linewidth = 1,
               color = "grey50",
               alpha = 0.75) +
  geom_text(aes(x = NMDS1, y = NMDS2, label = envvar),
            data = envfit_vecs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic() +
  theme(legend.position = "bottom")

# Ice plot
ice_facs <- filter(envfit_facs, str_detect(envvar, "ice_cat"))
ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = ice_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = ice_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

# Macrozooplankton plot
zoop_facs <- filter(envfit_facs, str_detect(envvar, "zoop"))
ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = zoop_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = zoop_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()
```

>>>>

library(cluster)
library(ggmosaic)
library(ggtree)
library(patchwork)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_stations %>%
  as_tibble() %>% 
  group_by(amlr.station, species) %>% 
  summarize(count_station = sum(count),
            .groups = "drop") %>% 
  group_by(species) %>%
  summarize(total_count = sum(count_station),
            stations_present = n(),
            station_frac = stations_present / total_stations) %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

frac_fn <- scales::label_percent(accuracy = 0.1)
pred_frac %>% 
  set_names(c("Species", "Ind", "Stations", "Fraction")) %>% 
  mutate(Fraction = frac_fn(Fraction)) %>% 
  knitr::kable(align = "lrr") %>% 
  kableExtra::kable_styling("striped",
                            full_width = FALSE) %>% 
  kableExtra::row_spec(cutoff_row, 
                       extra_css = "border-bottom: 2px dashed red")

@@

* Cluster 1 = open ocean? (ANFU)
* Cluster 2 = marginal ice zone (SNPT, ANPT, FUSE)
* Cluster 3 = pagophilic (ADPN, CRSE, LESE)

@@

pred_clust_sf <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(select(zoop_sf,
                   amlr.station,
                   geometry,
                   Year),
            by = "amlr.station") %>% 
  st_as_sf() %>% 
  st_transform(ant_proj())

@@

nmds_df <- sightings_nmds$points %>% 
  as_tibble(rownames = "amlr.station") %>% 
  rename_with(~ paste0("N", .x), starts_with("MDS")) %>% 
  right_join(pred_clust_sf %>% 
               as_tibble() %>% 
               select(amlr.station, pred_clust),
             by = "amlr.station")

@@

* Predator cluster associations with ice are unclear. May change after digging into Christian's ice data?
* Both predator cluster 1 (open-ish) and 2 (marginal ice zone) are associated with higher temperatures and macrozooplankton cluster 2 (small euphausiids, amphipods, etc)
* Predator cluster 3 (pagophilic) is associated with higher salinity, deeper mixed layer, and macrozooplankton cluster 3 (big euphausiids, Metridia, etc)
* Macrozooplankton cluster 1 (salps etc) evenly distributed across predator clusters

@@

# Hydrology plot
hydro_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               data = envfit_vecs,
               linewidth = 1,
               color = "grey50",
               alpha = 0.6) +
  geom_text(aes(x = NMDS1, y = NMDS2, label = envvar),
            data = envfit_vecs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

# Ice plot
ice_facs <- filter(envfit_facs, str_detect(envvar, "ice_cat"))
ice_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.6) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = ice_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = ice_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

# Macrozooplankton plot
zoop_facs <- filter(envfit_facs, str_detect(envvar, "zoop"))
zoop_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.6) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = zoop_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = zoop_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

layout <- "
AAAA
BBCC
"
hydro_plot + ice_plot + zoop_plot + 
  plot_layout(design = layout, guides = "collect")
```
